
---
title: "Automatic transmission versus manual tranmission based on MPG"
author: "Susana Arias Laso"
output: pdf_document
---

# Executive summary

The relationship between some of the variables in the mtcars dataset and the 
outcome, miles per gallon, of a set of regression models is explored by 
developing a regression analysis of the relevant statistical factors. More
specifically, the questions whether automatic or manual transmission is better
for mpg is addressed below.

```{r packages, echo=TRUE, results='hide'}
library("ggplot2")
require("cowplot")
```

The "am" variable in the mtcars dataset is of binary type, indicating manual
tranmission with "1" and automatic transmission with "0". The box plot can be 
seen in the appendix (Figure 1) indicating that manual transmission is more 
effective in terms of miles/gallon, the box plots indicate the media of mpg 
for the three number of cylinders in the dataset. With exception of 8-cylinder 
cars, which are independent of the type of transmission as the blue boxes 
indicate, one can say that manual cars are better in terms of miles/gallon.

Fit a linear model with transmission as the only variable, without taking into
consideration the number of cylinders or the weight of the cars.
```{r, echo=TRUE}
fit <- lm(mpg ~ am, data = mtcars)
coef <- summary(fit)$coefficients
coef
```
This model assumes one line through the data. As the slope of the linear model 
indicates, there is a `r coef[2, 1]` increase in miles/gallon going from 
automatic transmission to manual.

```{r, echo=TRUE}
confInterval <- confint(fit)
confInterval
```
The confidence interval indicates that with a 95% confidence, there is an
increase from `r confInterval[2, 1]` to `r confInterval[2, 2]` miles/gallon 
going from automatic transmission to manual.

# Multivariable regression analysis

An additional variable, the weight of the cars, in included in the analysis. In 
this way it is possible to model how other factors may interfere in the outcome
of this study, which is miles/gallon. See figure 2 in the appendix.

```{r, echo=TRUE}
fit2 <- lm(mpg ~ am + wt, data = mtcars)
coef2 <- summary(fit2)$coefficients
coef2
```

Adding a second variable to the model, the weight of the cars, namely, wt
one finds that the coefficient in front of the binary variable, am, is the 
change in the intercept between automatic and manual transmission. This model 
generates two parallel lines with different interceps and the same slope 
(`r coef2[3, 1]`), which indicates the change in mpg per unit increase 
(1000 lbs) in the car weight. The coefficient that multiplies the binary 
variable, `r coef2[2, 1]` represents the change in the intercept between 
automatic and manual transmission. However, having added the weight of the cars
as a second variable one can summarize that the transmission type is not really
relevant for fuel efficiency in comparison to the weight, as one can read from
the t-probability `r coef2[2, 4]`.

# Residual plots

The data sets corresponding to the two values of the binary variable are 
created, data1 and data0, to construct the regression lines with their 
residuals. The regression lines and residual plots are shown in Figure 3 in the
Appendix.

```{r, echo=TRUE}
y1 <- mtcars$mpg[mtcars$am == 1]
x1 <- mtcars$wt[mtcars$am == 1]
data1 <- data.frame(x1, y1)
fit1 <- lm(y1 ~ mtcars$am[mtcars$am == 1] + x1)
coef1 <- summary(fit1)$coefficients
```

```{r, echo=TRUE}
y0 <- mtcars$mpg[mtcars$am == 0]
x0 <- mtcars$wt[mtcars$am == 0]
data0 <- data.frame(x0, y0)
fit0 <- lm(y0 ~ mtcars$am[mtcars$am == 0] + x0)
coef0 <- summary(fit0)$coefficients
```


# Apendix

## This appendix contains supplementary figures and their respective R codes.

Box plot of the mpg(outcome) as a function of the type of transmission of the 
car. The number of cylinders was included as a second variable in the plot an
indicates that, except for 8-cylinder cars, it is not relevant to consider this
variable for a regression analysis.
```{r, echo=TRUE}
g <- ggplot(data = mtcars, aes(y = mpg, x = factor(am), fill = factor(cyl)))
g <- g + geom_boxplot()
g <- g + xlab("transmission") + ylab("miles/gallon")
g
```

This figure shows a point plot that indicates the direct influence the variable 
weight has on the mpg outcome. Therefore, this is a variable worth considering 
when analyzing whether automatic or manual transmission is better for mpg, given
that manual cars weight less than automatic ones.
```{r, echo=TRUE}
g <- ggplot(data = mtcars, aes(y = mpg, x = wt, color = factor(am)))
g <- g + geom_point() + xlab("weight (1000 lbs)") + ylab("miles/gallon")
g
```

The effect of the weight on the mpg variable is shown in these figures for the
two possible values of the binary variable, the transmission type.
```{r, echo=TRUE}
plot_manual <- ggplot(data1, aes(x = x1, y = y1)) + geom_point() + 
    geom_abline(intercept = coef1[1, 1], slope = coef1[2, 1]) + 
    stat_smooth(method = "lm", col = "blue") + 
    xlab("weight (1000 lbs)") + ylab("miles/gallon") + 
    ggtitle("manual")

plot_auto <- ggplot(data0, aes(x = x0, y = y0)) + geom_point() + 
    geom_abline(intercept = coef0[1, 1], slope = coef0[2, 1]) + 
    stat_smooth(method = "lm", col = "blue") + 
    xlab("weight (1000 lbs)") + ylab("") + 
    ggtitle("automatic")

plot_grid(plot_manual, plot_auto)
```




