
---
title: "Automatic transmission versus manual tranmission based on MPG"
author: "Susana Arias Laso"
output: pdf_document
---

# Executive summary

The relationship between some of the variables in the mtcars dataset and the 
outcome, miles per gallon, of a set of regression models is explored by 
developing a regression analysis of the relevant statistical factors. More
specifically, the questions whether automatic or manual transmission is better
for mpg is addressed below.

```{r packages, echo=TRUE, results='hide'}
library("ggplot2")
require("cowplot")
```

The "am" variable in the mtcars dataset is of binary type, indicating manual
tranmission with "1" and automatic transmission with "0". The following figure 
shows that manual transmission is more effective in terms of miles/gallon, the
box plots indicate the media of mpg for the three number of cylinders in the 
dataset. With exception of 8-cylinder cars, which are independent of the type 
of transmission as the blue boxes indicate, one can say that manual cars are 
better in terms of miles/gallon.

```{r, echo=TRUE}
g <- ggplot(data = mtcars, aes(y = mpg, x = factor(am), fill = factor(cyl)))
g <- g + geom_boxplot()
g <- g + xlab("transmission") + ylab("miles/gallon")
g
```

Fit a linear model with transmission as the only variable, without taking into
consideration the number of cylinders or the weight of the cars.
```{r, echo=TRUE}
fit <- lm(mpg ~ am, data = mtcars)
coef <- summary(fit)$coefficients
coef
```
This model assumes one line through the data. As the slope of the linear model 
indicates, there is a `r coef[2, 1]` increase in miles/gallon going from 
automatic transmission to manual.

# confidence interval

```{r, echo=TRUE}
confInterval <- confint(fit)
confInterval
```
The confidence interval indicates that with a 95% confidence, there is an
increase from `r confInterval[2, 1]` to `r confInterval[2, 2]` miles/gallon 
going from automatic transmission to manual.

# Multivariable regression analysis

```{r, echo=TRUE}
g <- ggplot(data = mtcars, aes(y = mpg, x = wt, color = factor(am)))
g <- g + geom_point() + xlab("weight (1000 lbs)") + ylab("miles/gallon")
g
```
This plot indicates that weight influences the mpg outcome directly, so this
is a variable worth having into account when analyzing whether automatic or
manual transmission is better for mpg, given that manual cars weight less than 
automatic cars

```{r, echo=TRUE}
fit2 <- lm(mpg ~ am + wt, data = mtcars)
coef2 <- summary(fit2)$coefficients
coef2
```

Adding a second variable to the model, the weight of the cars, namely, wt
one finds that the coefficient in front of the binary variable, am, is the 
change in the intercept between automatic and manual transmission. This model 
generates two parallel lines with different interceps and the same slope 
(`r coef2[3, 1]`), which indicates the change in mpg per unit increase 
(1000 lbs) in the car weight. The coefficient that multiplies the binary 
variable, `r coef2[2, 1]` represents the change in the intercept between 
automatic and manual transmission.


# Residual plots


```{r, echo=TRUE}
y1 <- mtcars$mpg[mtcars$am == 1]
x1 <- mtcars$wt[mtcars$am == 1]
data1 <- data.frame(x1, y1)
fit1 <- lm(y1 ~ mtcars$am[mtcars$am == 1] + x1)
coef1 <- summary(fit1)$coefficients
coef1
```


```{r, echo=TRUE}
y0 <- mtcars$mpg[mtcars$am == 0]
x0 <- mtcars$wt[mtcars$am == 0]
data0 <- data.frame(x0, y0)
fit0 <- lm(y0 ~ mtcars$am[mtcars$am == 0] + x0)
coef0 <- summary(fit0)$coefficients
coef0
```


# plot residuals of the two fits that correspond to the values of the binary
# variable (am), with their respective residuals
```{r, echo=TRUE}
e1 <- resid(fit1)
yhat1 <- predict(fit1)
```


```{r, echo=TRUE}
e0 <- resid(fit0)
yhat0 <- predict(fit0)
```


```{r, echo=TRUE}
plot_manual <- ggplot(data1, aes(x = x1, y = y1)) + geom_point() + 
    geom_abline(intercept = coef1[1, 1], slope = coef1[2, 1]) + 
    stat_smooth(method = "lm", col = "blue") + 
    xlab("weight (1000 lbs)") + ylab("miles/gallon") + 
    ggtitle("manual")

plot_auto <- ggplot(data0, aes(x = x0, y = y0)) + geom_point() + 
    geom_abline(intercept = coef0[1, 1], slope = coef0[2, 1]) + 
    stat_smooth(method = "lm", col = "blue") + 
    xlab("weight (1000 lbs)") + ylab("") + 
    ggtitle("automatic")

plot_grid(plot_manual, plot_auto)
```













